---
title: "Simmons Cohort DMP and DMR Analysis - Cell Deconvolution Ax (No ComBat), Continuous PM2.5"
author: "Gillian Goobie"
date: "2024_07_18"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    code_folding: hide
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    number_sections: yes
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE, echo=F}
library(tidyverse)
library(readxl)
library(sesame)
library(sesameData)
library(parallel)
library(here)
library(SummarizedExperiment)
library(knitr)
library(limma)
library(minfi)
library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
library(IlluminaHumanMethylationEPICmanifest)
library(RColorBrewer)
library(missMethyl)
library(minfiData)
library(Gviz)
library(DMRcate)
library(stringr)
library(writexl)
library(EpiDISH)
library(ChAMP)
library(doParallel)
library(psych)
library(lumi)
library(data.table)
library(qqman)
```

# Load Processed DNAm Data
"DNAmEPIC_Environment.RData" has all the other files as well as the original bVals and mVals prior to ComBat correction
```{r}
load("~/DNAmEPIC/DNAmEPIC_Environment.RData")
```

Below is loading the ComBat corrected bVals and mVals objects - not using these values in this analysis, which is not batch-corrected
```{r, eval=F}
load("~/DNAmEPIC/bVals_mVals_FullCohort_PostComBat_PlateArray.RData")
```


# Fix Simmons race based on chart review by Sharon (2023/11/01)
```{r}
#First create a dataframe of the patients with race that needs to be fixed. They're all White.
fix_race <- data.frame(
  ID=c(8354, 9434, 10526, 13285, 13310, 13619, 13714, 16924, 22220),
  new_race=rep("W", times=9),
  new_dich_Race=rep("White", times=9),
  new_d_Race=rep("White", times=9)
)

#Next perform the replacement using a left_join
meta <- meta %>% left_join(fix_race, by="ID") %>% 
  mutate(
    race=coalesce(new_race, race),
    dich_Race=coalesce(new_dich_Race, dich_Race)
  ) %>% 
  dplyr::select(-new_race, -new_dich_Race) #drop the temporary columns

targetsx <- targetsx %>% left_join(fix_race, by=c("ID.y"="ID")) %>% 
  mutate(
    race=coalesce(new_race, race),
    dich_Race=coalesce(new_dich_Race, dich_Race),
    d_Race=coalesce(new_d_Race, d_Race)
  ) %>% 
  dplyr::select(-new_race, -new_dich_Race, -new_d_Race) #drop the temporary columns
```

# ComBat Batch Correction
## Correct pd file
The pd file is otherwise known as the Sample_Sheet.csv or "targets" in the DNAmEPIC_Environment.RData. This is the file where the base data from the .idat files is retained that will allow us to look for batch effects.

First we need to define "Sample_Group" based on low vs high PM2.5 in 5yrs pre-sampling
```{r, eval=F}
targetsx$Sample_Group <- targetsx$PM5yrSamp_dich
```

```{r, eval=F}
#Need to first rename EPIC_ID to Sample_Name to be consistent with formatting required by ComBat
targets.c <- targetsx %>% rename("EPIC_ID"="Sample_Name", "cohort"="Project")
colnames(targets.c)
```

```{r, eval=F}
targets.c <- targets.c %>% dplyr::select(Sample_Name, Sample_Plate, Sample_Group, Pool_ID, Project, Sample_Well, Array, Slide, Basename)
targets.c$filename <- NA
targets.c$PM5yr_dich <- targets.c$Sample_Group
targets.c$Sample_Group <- 1
targets.c <- targets.c %>% mutate(across(c("Pool_ID", "Sample_Well", "Array", "filename"), as.factor))
str(targets.c)
```

```{r, eval=F}
targets.c$Sample_Plate <- droplevels(targets.c$Sample_Plate)
str(targets.c$Sample_Plate)
```


## Check Baseline SVD (Singular Variable Decomposition)
SVD will detect the number and nature of significant sources of variation in the methylation dataset and is helpful in determining if correction for batch effects is required.
```{r, eval=F}
bVals.df <- as.data.frame(bVals)
champ.SVD(beta=bVals.df, pd=targets.c)
```
This suggests that there are likely significant slide and plate batch effects here that should be corrected for.


## Batch Correct Plate and Array
```{r, eval=F}
mycombat.arrayplate <- champ.runCombat(beta=bVals.df, pd=targets.c, batchname=c("Array", "Sample_Plate"), logitTrans=TRUE)
```

```{r, eval=F}
mycombat.arrayplatedf <- as.data.frame(mycombat.arrayplate)
```

```{r, eval=F}
champ.SVD(beta=mycombat.arrayplatedf, pd=targets.c)
```



## Re-Make bVals Dataframe
```{r, eval=F}
bVals <- mycombat.arrayplate
```

```{r, eval=F}
#Convert bVals matrix to mVals matrix 
mVals <- beta2m(bVals)
```

```{r, eval=F}
rm(contMatrix, DMPs, fit, fit2, fix_race, targets.c)
```

## Save bVals and mVals to R object
```{r, eval=F}
save(bVals, mVals, file = "bVals_mVals_FullCohort_PostComBat_PlateArray.RData")
```


# Get rid of CARE-PF samples
First in targetsx and meta
```{r}
targetsx <- targetsx %>% filter(cohort=="Simmons")
meta <- meta %>% filter(cohort=="Simmons")
```
Takes us down to 306 observations.

```{r}
dim(mVals)
Simmons_IDs <- targetsx$ID.x
mVals <- mVals[,Simmons_IDs]
dim(mVals)
```

```{r}
dim(bVals)
bVals <- bVals[,Simmons_IDs]
dim(bVals)
```

# Making New Dichotomized Cutpoints for PM2.5
## targetsx - Make Dichotomized PM2.5
### 5yrs Pre-Sampling
Will use the Simmons median for these analyses as the previous cutpoints that were shared between the two cohorts in the full cohort analyses often leave 0 patients from Simmons in the high exposure group and 0 patients from Simmons in the low exposure group.
```{r}
summary(targetsx$PM_5yrPreSamp)
targetsx$PM5yrSamp_dich <- cut(targetsx$PM_5yrPreSamp,
                      breaks=c(0, 12.07, 100),
                      labels=c("Low", "High"))
summary(targetsx$PM5yrSamp_dich)
class(targetsx$PM5yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$PM_3moPreSamp)
targetsx$PM3moSamp_dich <- cut(targetsx$PM_3moPreSamp,
                      breaks=c(0, 10.50, 100),
                      labels=c("Low", "High"))
summary(targetsx$PM3moSamp_dich)
class(targetsx$PM3moSamp_dich)
```


# Make New Dichotomized Cut-points for Constituents
## targetsx - Make Dichotomized SO4
### 5yrs Pre-Sampling
```{r}
summary(targetsx$SO4_5yrPreSamp)
targetsx$SO45yrSamp_dich <- cut(targetsx$SO4_5yrPreSamp,
                      breaks=c(0, 4.17, 100),
                      labels=c("Low", "High"))
summary(targetsx$SO45yrSamp_dich)
class(targetsx$SO45yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$SO4_3moPreSamp)
targetsx$SO43moSamp_dich <- cut(targetsx$SO4_3moPreSamp,
                      breaks=c(0, 2.95, 100),
                      labels=c("Low", "High"))
summary(targetsx$SO43moSamp_dich)
class(targetsx$SO43moSamp_dich)
```



## targetsx - Make Dichotomized NO3
### 5yrs Pre-Sampling
```{r}
summary(targetsx$NO3_5yrPreSamp)
targetsx$NO35yrSamp_dich <- cut(targetsx$NO3_5yrPreSamp,
                      breaks=c(0, 1.11, 100),
                      labels=c("Low", "High"))
summary(targetsx$NO35yrSamp_dich)
class(targetsx$NO35yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$NO3_3moPreSamp)
targetsx$NO33moSamp_dich <- cut(targetsx$NO3_3moPreSamp,
                      breaks=c(0, 1.07, 100),
                      labels=c("Low", "High"))
summary(targetsx$NO33moSamp_dich)
class(targetsx$NO33moSamp_dich)
```


## targetsx - Make Dichotomized NH4
### 5yrs Pre-Sampling
```{r}
summary(targetsx$NH4_5yrPreSamp)
targetsx$NH45yrSamp_dich <- cut(targetsx$NH4_5yrPreSamp,
                      breaks=c(0, 1.61, 100),
                      labels=c("Low", "High"))
summary(targetsx$NH45yrSamp_dich)
class(targetsx$NH45yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$NH4_3moPreSamp)
targetsx$NH43moSamp_dich <- cut(targetsx$NH4_3moPreSamp,
                      breaks=c(0, 1.33, 100),
                      labels=c("Low", "High"))
summary(targetsx$NH43moSamp_dich)
class(targetsx$NH43moSamp_dich)
```


## targetsx - Make Dichotomized BC
### 5yrs Pre-Sampling
```{r}
summary(targetsx$BC_5yrPreSamp)
targetsx$BC5yrSamp_dich <- cut(targetsx$BC_5yrPreSamp,
                      breaks=c(0, 0.85, 100),
                      labels=c("Low", "High"))
summary(targetsx$BC5yrSamp_dich)
class(targetsx$BC5yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$BC_3moPreSamp)
targetsx$BC3moSamp_dich <- cut(targetsx$BC_3moPreSamp,
                      breaks=c(0, 0.83, 100),
                      labels=c("Low", "High"))
summary(targetsx$BC3moSamp_dich)
class(targetsx$BC3moSamp_dich)
```


## targetsx - Make Dichotomized OM
### 5yrs Pre-Sampling
```{r}
summary(targetsx$OM_5yrPreSamp)
targetsx$OM5yrSamp_dich <- cut(targetsx$OM_5yrPreSamp,
                      breaks=c(0, 3.30, 100),
                      labels=c("Low", "High"))
summary(targetsx$OM5yrSamp_dich)
class(targetsx$OM5yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$OM_3moPreSamp)
targetsx$OM3moSamp_dich <- cut(targetsx$OM_3moPreSamp,
                      breaks=c(0, 3.20, 100),
                      labels=c("Low", "High"))
summary(targetsx$OM3moSamp_dich)
class(targetsx$OM3moSamp_dich)
```


## targetsx - Make Dichotomized SS
### 5yrs Pre-Sampling
```{r}
summary(targetsx$SS_5yrPreSamp)
targetsx$SS5yrSamp_dich <- cut(targetsx$SS_5yrPreSamp,
                      breaks=c(0, 0.19, 100),
                      labels=c("Low", "High"))
summary(targetsx$SS5yrSamp_dich)
class(targetsx$SS5yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$SS_3moPreSamp)
targetsx$SS3moSamp_dich <- cut(targetsx$SS_3moPreSamp,
                      breaks=c(0, 0.17, 100),
                      labels=c("Low", "High"))
summary(targetsx$SS3moSamp_dich)
class(targetsx$SS3moSamp_dich)
```


## targetsx - Make Dichotomized Soil
### 5yrs Pre-Sampling
```{r}
summary(targetsx$Soil_5yrPreSamp)
targetsx$Soil5yrSamp_dich <- cut(targetsx$Soil_5yrPreSamp,
                      breaks=c(0, 0.51, 100),
                      labels=c("Low", "High"))
summary(targetsx$Soil5yrSamp_dich)
class(targetsx$Soil5yrSamp_dich)
```


### 3mo Pre-Sampling
```{r}
summary(targetsx$Soil_3moPreSamp)
targetsx$Soil3moSamp_dich <- cut(targetsx$Soil_3moPreSamp,
                      breaks=c(0, 0.43, 100),
                      labels=c("Low", "High"))
summary(targetsx$Soil3moSamp_dich)
class(targetsx$Soil3moSamp_dich)
```


# Releveling factor variables in targetsx
```{r}
targetsx$EPIC_ID <- as.factor(targetsx$EPIC_ID)
targetsx$sex <- fct_relevel(targetsx$sex, c("M","F"))
targetsx$cohort <- fct_relevel(targetsx$cohort, c("Simmons"))
targetsx$race <- fct_relevel(targetsx$race, c("W","B","A","U"))
targetsx$dich_Race <- fct_relevel(targetsx$dich_Race, c("White","Non-White"))
targetsx$smokeHx <- fct_relevel(targetsx$smokeHx, c("Never","Former","Always","Unknown"))

targetsx$dx <- fct_relevel(targetsx$dx, c("IPF"))
targetsx$dx_group <- fct_relevel(targetsx$dx_group, c("IPF"))
targetsx$dx_IPF <- fct_relevel(targetsx$dx_IPF, c("IPF"))
targetsx$PM5yrSamp_dich <- fct_relevel(targetsx$PM5yrSamp_dich, c("Low"))
targetsx$PM1yrSamp_dich <- fct_relevel(targetsx$PM1yrSamp_dich, c("Low"))
targetsx$PM6moSamp_dich <- fct_relevel(targetsx$PM6moSamp_dich, c("Low"))
targetsx$PM3moSamp_dich <- fct_relevel(targetsx$PM3moSamp_dich, c("Low"))
targetsx$PM1moSamp_dich <- fct_relevel(targetsx$PM1moSamp_dich, c("Low"))
targetsx$SO45yrSamp_dich <- fct_relevel(targetsx$SO45yrSamp_dich, c("Low"))
targetsx$SO41yrSamp_dich <- fct_relevel(targetsx$SO41yrSamp_dich, c("Low"))
targetsx$SO46moSamp_dich <- fct_relevel(targetsx$SO46moSamp_dich, c("Low"))
targetsx$SO43moSamp_dich <- fct_relevel(targetsx$SO43moSamp_dich, c("Low"))
targetsx$SO41moSamp_dich <- fct_relevel(targetsx$SO41moSamp_dich, c("Low"))
targetsx$NO35yrSamp_dich <- fct_relevel(targetsx$NO35yrSamp_dich, c("Low"))
targetsx$NO31yrSamp_dich <- fct_relevel(targetsx$NO31yrSamp_dich, c("Low"))
targetsx$NO36moSamp_dich <- fct_relevel(targetsx$NO36moSamp_dich, c("Low"))
targetsx$NO33moSamp_dich <- fct_relevel(targetsx$NO33moSamp_dich, c("Low"))
targetsx$NO31moSamp_dich <- fct_relevel(targetsx$NO31moSamp_dich, c("Low"))
targetsx$NH45yrSamp_dich <- fct_relevel(targetsx$NH45yrSamp_dich, c("Low"))
targetsx$NH41yrSamp_dich <- fct_relevel(targetsx$NH41yrSamp_dich, c("Low"))
targetsx$NH46moSamp_dich <- fct_relevel(targetsx$NH46moSamp_dich, c("Low"))
targetsx$NH43moSamp_dich <- fct_relevel(targetsx$NH43moSamp_dich, c("Low"))
targetsx$NH41moSamp_dich <- fct_relevel(targetsx$NH41moSamp_dich, c("Low"))
targetsx$BC5yrSamp_dich <- fct_relevel(targetsx$BC5yrSamp_dich, c("Low"))
targetsx$BC1yrSamp_dich <- fct_relevel(targetsx$BC1yrSamp_dich, c("Low"))
targetsx$BC6moSamp_dich <- fct_relevel(targetsx$BC6moSamp_dich, c("Low"))
targetsx$BC3moSamp_dich <- fct_relevel(targetsx$BC3moSamp_dich, c("Low"))
targetsx$BC1moSamp_dich <- fct_relevel(targetsx$BC1moSamp_dich, c("Low"))
targetsx$OM5yrSamp_dich <- fct_relevel(targetsx$OM5yrSamp_dich, c("Low"))
targetsx$OM1yrSamp_dich <- fct_relevel(targetsx$OM1yrSamp_dich, c("Low"))
targetsx$OM6moSamp_dich <- fct_relevel(targetsx$OM6moSamp_dich, c("Low"))
targetsx$OM3moSamp_dich <- fct_relevel(targetsx$OM3moSamp_dich, c("Low"))
targetsx$OM1moSamp_dich <- fct_relevel(targetsx$OM1moSamp_dich, c("Low"))
targetsx$SS5yrSamp_dich <- fct_relevel(targetsx$SS5yrSamp_dich, c("Low"))
targetsx$SS1yrSamp_dich <- fct_relevel(targetsx$SS1yrSamp_dich, c("Low"))
targetsx$SS6moSamp_dich <- fct_relevel(targetsx$SS6moSamp_dich, c("Low"))
targetsx$SS3moSamp_dich <- fct_relevel(targetsx$SS3moSamp_dich, c("Low"))
targetsx$SS1moSamp_dich <- fct_relevel(targetsx$SS1moSamp_dich, c("Low"))
targetsx$Soil5yrSamp_dich <- fct_relevel(targetsx$Soil5yrSamp_dich, c("Low"))
targetsx$Soil1yrSamp_dich <- fct_relevel(targetsx$Soil1yrSamp_dich, c("Low"))
targetsx$Soil6moSamp_dich <- fct_relevel(targetsx$Soil6moSamp_dich, c("Low"))
targetsx$Soil3moSamp_dich <- fct_relevel(targetsx$Soil3moSamp_dich, c("Low"))
targetsx$Soil1moSamp_dich <- fct_relevel(targetsx$Soil1moSamp_dich, c("Low"))
str(targetsx)
```

## Convert Date Columns to Date Format
Next I need to convert all date columns to proper format
```{r}
targetsx <- targetsx %>% 
  mutate_at(c("sample_date", "PM_date"), as.Date)
```

# Cell-Type Deconvolution
```{r}
out.l <- epidish(beta.m = bVals, ref.m = centEpiFibIC.m, method = "RPC")
out.l$estF
dim(out.l$ref)
dim(out.l$dataREF)
```
Can see that the fast majority of samples had nearly 100% immune cells (IC)

```{r}
BloodFrac.m <- epidish(beta.m = bVals, ref.m = centDHSbloodDMC.m, method = "RPC")$estF
```

```{r}
boxplot(BloodFrac.m)
```
Neutrophils are clearly the most highly represented cell type.

## Refine Cell Distributions into Principal Components
Need to ensure that none of our columns have values !=0
```{r}
colSums(is.na(BloodFrac.m))
# All Eosino are =0 after batch correction, so will remove this column from the datafram
BloodFrac.m <- as.data.frame(BloodFrac.m)
BloodFrac.m <- BloodFrac.m %>% dplyr::select(!Eosino)
```

Need to normalize all the cell types so they are on the same scale
```{r}
BloodFrac_norm <- scale(BloodFrac.m)
head(BloodFrac.m)
head(BloodFrac_norm)
```

Make a correlation matrix
```{r}
library(ggcorrplot)
corr_matrix <- cor(BloodFrac_norm)
ggcorrplot(corr_matrix)
```

Plot scatterplot of CD4T vs Neuts for simple visualization of PCs
```{r}
BloodFrac.mx <- as.data.frame(BloodFrac.m)

(BloodFrac.mx %>% ggplot(aes(x=Neutro, y=CD4T))+
   geom_point()+
   labs(x="Neutrophil Proportion", y="CD4+ T-Cell Proportion")+
   theme(plot.title = element_text(hjust = 0.5)))
```


```{r}
BloodFrac_pca <- princomp(corr_matrix)
summary(BloodFrac_pca)
```

```{r}
BloodFrac_pca$loadings[,1:5]
```

```{r}
library(factoextra)
fviz_eig(BloodFrac_pca, addlabels=T)
```
This scree plot shows that 54% of the variation in cell composition is retained in PC1, then the next 18.5% in PC2, and so on...

Another method of scree plot
```{r}
screeplot(BloodFrac_pca, type="l", main="Scree Plot for Cell Distributions")
```
Basically the same as above, just with different y-axis scale

```{r}
fviz_pca_var(BloodFrac_pca, col.var = "black")
```
Can see that Neutrophils are a major driver of the PCA that moves in the opposite direction of other cell types. 
Lines that group together are positively correlated with each other, so CD4 T-cells and B-cells are positively correlated, as are Eos/NK/CD8 T-cells. CD4 T-Cells and CD8 T-cells are negatively correlated with each other

```{r}
fviz_cos2(BloodFrac_pca, choice = "var", axes = 1:5)
```
This shows that Neuts are contributing the most the the first 5 PCAs, whereas Eos are contributing the least.

```{r}
BloodFrac_norm <- as.data.frame(BloodFrac_norm)
BloodFrac_PCs <- principal(BloodFrac_norm, nfactors=5, rotate="none")
BloodFrac_PCs
```
The cumulative variance after PC5 is 0.94, suggesting that by using the first 5 PCs, we are only missing around 6% of the variance in our cell distribution dataset.

```{r}
BloodFrac_PC.fit <- prcomp(~ B + NK + CD4T + CD8T + Mono + Neutro,
                           data=BloodFrac_norm, 
                           scale=TRUE)
summary(BloodFrac_PC.fit)
```

Then convert this into dataframe where there is a PC assigned to each row (i.e. each individual)
```{r}
new_BloodFrac <- as.data.frame(predict(BloodFrac_PC.fit, BloodFrac_norm)[,1:5])
```

## Add Immune Cell Distribution Columns to targetsx
```{r}
new_BloodFrac <- as.data.frame(new_BloodFrac) 
EPIC_ID <- as.data.frame(Simmons_IDs)
EPIC_ID <- sub("Simmons.","",EPIC_ID$Simmons_IDs)
EPIC_ID <- as.data.frame(EPIC_ID) 
```

```{r}
new_BloodFrac <- cbind(EPIC_ID, new_BloodFrac)
```

```{r}
targetsx <- left_join(targetsx, new_BloodFrac, by="EPIC_ID")
```


# PM2.5 Differential Methylation
## Differential methylation by Continuous PM2.5 in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by PM_5yrPreSamp
```{r}
# Convert PM_5yrPreSamp to numeric
PM5yrPreSamp_cont <- as.numeric(targetsx$PM_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ PM5yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there is 0 CpGs with lower DNAm levels and 2 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousPM5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.57 for this analysis.



## Differential methylation by Continuous PM2.5 in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert PM_5yrPreSamp to numeric
PM5yrPreSamp_cont <- as.numeric(targetsx$PM_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ PM5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 530 CpGs with lower DNAm levels in low vs high PM2.5 in 5yrs pre-sampling and 337 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousPM5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.19 for this analysis.



## Differential methylation by Continuous PM2.5 in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert PM_5yrPreSamp to numeric
PM5yrPreSamp_cont <- as.numeric(targetsx$PM_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ PM5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 661 CpGs with lower DNAm levels and 500 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```


Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousPM5yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.30 for this analysis.



### PM5yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=-log10(8.97e-5),
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```

```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
PM5yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in PM5yr_manhattan to match the function's expected names
names(PM5yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
PM5yr_manhattan$CHR <- as.numeric(sub("chr", "", PM5yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified PM5yr_manhattan dataframe
gg.manhattan(PM5yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```



QQ plot
```{r}
qq(DMPs$P.Value)
```



### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "PM5yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```


```{r}
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
results.ranges_PM5yr_adj <- extractRanges(DMRs)
results.ranges_PM5yr_adj
```


### Gene Ontology Adjusted Analysis
```{r}
enrichment_GO_PM5yr_adj <- goregion(results.ranges_PM5yr_adj[1:168], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
enrichment_GO_PM5yr_adj <- enrichment_GO_PM5yr_adj[order(enrichment_GO_PM5yr_adj$P.DE),]
head(as.matrix(enrichment_GO_PM5yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
results.ranges_PM5yr_adj <- as.data.frame(results.ranges_PM5yr_adj)

write.table(results.ranges_PM5yr_adj, file="Simmons_ContinuousPM5yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Remove unnecessary files
```{r}
rm(list=ls(pattern="^enrichment"))
rm(list=ls(pattern="^BloodFrac"))
rm(list=ls(pattern="^fit"))
rm(list=ls(pattern="^results"))
rm(list=c("meta", "EPIC_ID", "dat", "new_BloodFrac"))
```



# SO4 Differential Methylation
## Differential methylation by Continuous SO4 in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by SO4_5yrPreSamp
```{r}
# Convert SO4_5yrPreSamp to numeric
SO45yrPreSamp_cont <- as.numeric(targetsx$SO4_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ SO45yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 2557 CpGs with lower DNAm levels and 2247 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousSO45yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.62 for this analysis.



## Differential methylation by Continuous SO4 in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert SO4_5yrPreSamp to numeric
SO45yrPreSamp_cont <- as.numeric(targetsx$SO4_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ SO45yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 1545 CpGs with lower DNAm levels in low vs high SO4 in 5yrs pre-sampling and 1094 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousSO45yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.57 for this analysis.



## Differential methylation by Continuous SO4 in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert SO4_5yrPreSamp to numeric
SO45yrPreSamp_cont <- as.numeric(targetsx$SO4_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ SO45yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 1426 CpGs with lower DNAm levels and 968 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousSO45yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.49 for this analysis.

### SO45yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=-log10(1.846e-04),
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```



```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
SO45yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in SO45yr_manhattan to match the function's expected names
names(SO45yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
SO45yr_manhattan$CHR <- as.numeric(sub("chr", "", SO45yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified PM5yr_manhattan dataframe
gg.manhattan(SO45yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```



QQ plot
```{r}
qq(DMPs$P.Value)
```




### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "SO45yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```


```{r}
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
results.ranges_SO45yr_adj <- extractRanges(DMRs)
results.ranges_SO45yr_adj
```



### Gene Ontology Adjusted Analysis
```{r}
enrichment_GO_SO45yr_adj <- goregion(results.ranges_SO45yr_adj[1:334], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
enrichment_GO_SO45yr_adj <- enrichment_GO_SO45yr_adj[order(enrichment_GO_SO45yr_adj$P.DE),]
head(as.matrix(enrichment_GO_SO45yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
results.ranges_SO45yr_adj <- as.data.frame(results.ranges_SO45yr_adj)

write.table(results.ranges_SO45yr_adj, file="Simmons_ContinuousSO45yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```



# NO3 Differential Methylation
## Differential methylation by Continuous NO3 in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by NO3_5yrPreSamp
```{r}
# Convert NO3_5yrPreSamp to numeric
NO35yrPreSamp_cont <- as.numeric(targetsx$NO3_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ NO35yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousNO35yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 0.92 for this analysis.



## Differential methylation by Continuous NO3 in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert NO3_5yrPreSamp to numeric
NO35yrPreSamp_cont <- as.numeric(targetsx$NO3_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ NO35yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousNO35yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 0.99 for this analysis.



## Differential methylation by Continuous NO3 in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert NO3_5yrPreSamp to numeric
NO35yrPreSamp_cont <- as.numeric(targetsx$NO3_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ NO35yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousNO35yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 0.92 for this analysis.



### NO35yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=NULL,
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```

```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
NO35yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in NO35yr_manhattan to match the function's expected names
names(NO35yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
NO35yr_manhattan$CHR <- as.numeric(sub("chr", "", NO35yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified NO35yr_manhattan dataframe
gg.manhattan(NO35yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```



QQ plot
```{r}
qq(DMPs$P.Value)
```



### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "NO35yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```

Can't run as no sig CpGs
```{r}
#DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
#results.ranges_NO35yr_adj <- extractRanges(DMRs)
#results.ranges_NO35yr_adj
```


### Gene Ontology Adjusted Analysis
```{r}
#enrichment_GO_NO35yr_adj <- goregion(results.ranges_NO35yr_adj[1:6], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
#enrichment_GO_NO35yr_adj <- enrichment_GO_NO35yr_adj[order(enrichment_GO_NO35yr_adj$P.DE),]
#head(as.matrix(enrichment_GO_NO35yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
#results.ranges_NO35yr_adj <- as.data.frame(results.ranges_NO35yr_adj)

#write.table(results.ranges_NO35yr_adj, file="Simmons_ContinuousNO35yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```



# NH4 Differential Methylation
## Differential methylation by Continuous NH4 in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by NH4_5yrPreSamp
```{r}
# Convert NH4_5yrPreSamp to numeric
NH45yrPreSamp_cont <- as.numeric(targetsx$NH4_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ NH45yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 2361 CpGs with lower DNAm levels and 2090 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousNH45yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.61 for this analysis.



## Differential methylation by Continuous NH4 in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert NH4_5yrPreSamp to numeric
NH45yrPreSamp_cont <- as.numeric(targetsx$NH4_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ NH45yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 954 CpGs with lower DNAm levels and 645 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousNH45yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.44 for this analysis.



## Differential methylation by Continuous NH4 in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert NH4_5yrPreSamp to numeric
NH45yrPreSamp_cont <- as.numeric(targetsx$NH4_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ NH45yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 1018 CpGs with lower DNAm levels and 779 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousNH45yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.47 for this analysis.


### NH45yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=-log10(1.39e-04),
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```



```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
NH45yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in NH45yr_manhattan to match the function's expected names
names(NH45yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
NH45yr_manhattan$CHR <- as.numeric(sub("chr", "", NH45yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified NH45yr_manhattan dataframe
gg.manhattan(NH45yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```



QQ plot
```{r}
qq(DMPs$P.Value)
```



### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "NH45yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```


```{r}
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
results.ranges_NH45yr_adj <- extractRanges(DMRs)
results.ranges_NH45yr_adj
```



### Gene Ontology Adjusted Analysis
```{r}
enrichment_GO_NH45yr_adj <- goregion(results.ranges_NH45yr_adj[1:261], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
enrichment_GO_NH45yr_adj <- enrichment_GO_NH45yr_adj[order(enrichment_GO_NH45yr_adj$P.DE),]
head(as.matrix(enrichment_GO_NH45yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
results.ranges_NH45yr_adj <- as.data.frame(results.ranges_NH45yr_adj)

write.table(results.ranges_NH45yr_adj, file="Simmons_ContinuousNH45yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```



# BC Differential Methylation
## Differential methylation by Continuous BC in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by BC_5yrPreSamp
```{r}
# Convert BC_5yrPreSamp to numeric
BC5yrPreSamp_cont <- as.numeric(targetsx$BC_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ BC5yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousBC5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.66 for this analysis.



## Differential methylation by Continuous BC in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert BC_5yrPreSamp to numeric
BC5yrPreSamp_cont <- as.numeric(targetsx$BC_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ BC5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousBC5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.55 for this analysis.



## Differential methylation by Continuous BC in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert BC_5yrPreSamp to numeric
BC5yrPreSamp_cont <- as.numeric(targetsx$BC_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ BC5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousBC5yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.63 for this analysis.



### BC5yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=NULL,
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```

```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
BC5yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in BC5yr_manhattan to match the function's expected names
names(BC5yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
BC5yr_manhattan$CHR <- as.numeric(sub("chr", "", BC5yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified BC5yr_manhattan dataframe
gg.manhattan(BC5yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```




QQ plot
```{r}
qq(DMPs$P.Value)
```



### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "BC5yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```

Can't run as no sig CpGs
```{r}
#DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
#results.ranges_BC5yr_adj <- extractRanges(DMRs)
#results.ranges_BC5yr_adj
```


### Gene Ontology Adjusted Analysis
```{r}
#enrichment_GO_BC5yr_adj <- goregion(results.ranges_BC5yr_adj[1:6], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
#enrichment_GO_BC5yr_adj <- enrichment_GO_BC5yr_adj[order(enrichment_GO_BC5yr_adj$P.DE),]
#head(as.matrix(enrichment_GO_BC5yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
#results.ranges_BC5yr_adj <- as.data.frame(results.ranges_BC5yr_adj)

#write.table(results.ranges_BC5yr_adj, file="Simmons_ContinuousBC5yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```



# OM Differential Methylation
## Differential methylation by Continuous OM in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by OM_5yrPreSamp
```{r}
# Convert OM_5yrPreSamp to numeric
OM5yrPreSamp_cont <- as.numeric(targetsx$OM_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ OM5yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 2 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousOM5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 0.87 for this analysis.



## Differential methylation by Continuous OM in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert OM_5yrPreSamp to numeric
OM5yrPreSamp_cont <- as.numeric(targetsx$OM_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ OM5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousOM5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 0.87 for this analysis.



## Differential methylation by Continuous OM in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert OM_5yrPreSamp to numeric
OM5yrPreSamp_cont <- as.numeric(targetsx$OM_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ OM5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousOM5yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.00 for this analysis.


### OM5yr Manhattan Plot
```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
OM5yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in OM5yr_manhattan to match the function's expected names
names(OM5yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
OM5yr_manhattan$CHR <- as.numeric(sub("chr", "", OM5yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified OM5yr_manhattan dataframe
gg.manhattan(OM5yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```


QQ plot
```{r}
qq(DMPs$P.Value)
```



### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "OM5yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```

No sig CpGs so can't run DMR or GO
```{r}
#DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
#results.ranges_OM5yr_adj <- extractRanges(DMRs)
#results.ranges_OM5yr_adj
```


### Gene Ontology Adjusted Analysis
```{r}
#enrichment_GO_OM5yr_adj <- goregion(results.ranges_OM5yr_adj[1:6], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
#enrichment_GO_OM5yr_adj <- enrichment_GO_OM5yr_adj[order(enrichment_GO_OM5yr_adj$P.DE),]
#head(as.matrix(enrichment_GO_OM5yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
#results.ranges_OM5yr_adj <- as.data.frame(results.ranges_OM5yr_adj)

#write.table(results.ranges_OM5yr_adj, file="Simmons_ContinuousOM5yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```



# SS Differential Methylation
## Differential methylation by Continuous SS in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by SS_5yrPreSamp
```{r}
# Convert SS_5yrPreSamp to numeric
SS5yrPreSamp_cont <- as.numeric(targetsx$SS_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ SS5yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 1138 CpGs with lower DNAm levels and 915 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousSS5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.85 for this analysis.



## Differential methylation by Continuous SS in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert SS_5yrPreSamp to numeric
SS5yrPreSamp_cont <- as.numeric(targetsx$SS_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ SS5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 526 CpGs with lower DNAm levels in low vs high SS in 5yrs pre-sampling and 276 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousSS5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.64 for this analysis.



## Differential methylation by Continuous SS in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert SS_5yrPreSamp to numeric
SS5yrPreSamp_cont <- as.numeric(targetsx$SS_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ SS5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 628 CpGs with lower DNAm levels and 346 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousSS5yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.55 for this analysis.


### SS5yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=-log10(7.5e-05),
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```


```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
SS5yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in SS5yr_manhattan to match the function's expected names
names(SS5yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
SS5yr_manhattan$CHR <- as.numeric(sub("chr", "", SS5yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified SS5yr_manhattan dataframe
gg.manhattan(SS5yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```


QQ plot
```{r}
qq(DMPs$P.Value)
```


### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "SS5yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```


```{r}
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
results.ranges_SS5yr_adj <- extractRanges(DMRs)
results.ranges_SS5yr_adj
```


### Gene Ontology Adjusted Analysis
```{r}
enrichment_GO_SS5yr_adj <- goregion(results.ranges_SS5yr_adj[1:140], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
enrichment_GO_SS5yr_adj <- enrichment_GO_SS5yr_adj[order(enrichment_GO_SS5yr_adj$P.DE),]
head(as.matrix(enrichment_GO_SS5yr_adj), 10)
```

### Write DMR Results to .csv
```{r}
results.ranges_SS5yr_adj <- as.data.frame(results.ranges_SS5yr_adj)

write.table(results.ranges_SS5yr_adj, file="Simmons_ContinuousSS5yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```



# Soil Differential Methylation
## Differential methylation by Continuous Soil in 5yrs Pre-Sampling - Unadjusted
First analysis I will do is evaluate for differential methylation by Soil_5yrPreSamp
```{r}
# Convert Soil_5yrPreSamp to numeric
Soil5yrPreSamp_cont <- as.numeric(targetsx$Soil_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ Soil5yrPreSamp_cont, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsUnadjusted_ContinuousSoil5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.21 for this analysis.



## Differential methylation by Continuous Soil in 5yrs Pre-Sampling - Adjusted
```{r}
# Convert Soil_5yrPreSamp to numeric
Soil5yrPreSamp_cont <- as.numeric(targetsx$Soil_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ Soil5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 0 CpGs with lower DNAm levels in low vs high Soil in 5yrs pre-sampling and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousSoil5yrPreSamp_DMPs_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.30 for this analysis.



## Differential methylation by Continuous Soil in 5yrs Pre-Sampling - Covariate + Cell Type Adjustment
```{r}
# Convert Soil_5yrPreSamp to numeric
Soil5yrPreSamp_cont <- as.numeric(targetsx$Soil_5yrPreSamp)

# Create a design matrix including the continuous variable
design <- model.matrix(~ Soil5yrPreSamp_cont+sex+age_dx+d_Race+smokeHx+dx_IPF+PC1+PC2+PC3+PC4+PC5, data=targetsx)

# Fit the linear model
fit <- lmFit(mVals, design)

# Apply eBayes
fit2 <- eBayes(fit)
```

```{r}
# look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
So we can see here that there are 1 CpGs with lower DNAm levels and 0 CpGs with higher DNAm levels


```{r}
DMPs <- topTable(fit2, num=Inf, coef=2, genelist=annoSub)
head(DMPs)
```

Calculate mean P-value for DMP analysis
```{r}
mean(DMPs$P.Value)
```

Write the DMPs df to a table
```{r, eval=F}
write.table(DMPs, file="SimmonsAdjusted_ContinuousSoil5yrPreSamp_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```

Calculating the Genomic Inflation Factor
```{r}
DMPs$chisq <- qchisq(1-DMPs$P.Value,1)
gif <- median(DMPs$chisq)/qchisq(0.5,1)
gif
```
The genomic inflation factor (lambda) is 1.19 for this analysis.

### Soil5yr Manhattan Plot
```{r}
gg.manhattan = function(dataframe, title=NULL, cols = c('red', 'skyblue'),
                        chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'rsid', gene.fld=NULL,
                        chrs = 1:22, ymax = NULL,
                        suggestiveline=0, genomewideline=-log10(1e-07),
                        annotate=F, annotate.fld=NULL, SNPlist=NULL,lab2lim=NULL,lab3lim=NULL) {
  library(data.table)
  library(ggplot2)
  library(dplyr)
  library(ggrepel)

  if (annotate & is.null(SNPlist))
    stop("You requested annotation but provided no SNPlist!")
  #if(annotate & any(! unlist(SNPlist) %in% dataframe[[snp.fld]]))
  #  stop('No SNPs in SNPlist are found in the dataframe')

  if(is.data.table(dataframe)) dataframe <- as.data.frame(dataframe)

  stopifnot(all(c(chr.fld, pos.fld, p.fld, snp.fld) %in% names(dataframe)),
            is.data.frame(dataframe),
            all(!is.na(dataframe[[chr.fld]])),
            all(unlist(lapply(dataframe[c(chr.fld, pos.fld, p.fld)], is.numeric))))
  if(is.null(gene.fld)){
    d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld)]
    names(d) <- c('CHR', 'BP', 'P', 'SNP')
  }else{
  d <- dataframe[c(chr.fld, pos.fld, p.fld, snp.fld, gene.fld)]
  names(d) <- c('CHR', 'BP', 'P', 'SNP','Gene')
  }

  #limit to only chrs 1-22
  d <- d %>% filter(CHR %in% chrs, P > 0, P <= 1, !is.na(BP)) %>%
    mutate(logp = -log10(P)) %>% arrange(CHR, BP)

  # get position references for each CHR:
  d$CHR <- factor(d$CHR)
  posDat <- d %>% group_by(CHR) %>% dplyr::summarise(maxBP = max(BP))

  tmp <- data.frame(0, 0); names(tmp) <- names(posDat)
  posDat <- rbind(tmp, posDat)
  posDat$cumBP <- sapply(1:nrow(posDat), function(x) sum(posDat$maxBP[1:x]))
  posDat <- data.frame(CHR = posDat$CHR[-1], addBP = posDat$cumBP[-nrow(posDat)])

  d <- merge(d, posDat, all.x = T)

  d <- d %>% mutate(pos = BP + addBP)
  ticks <- d %>% group_by(CHR) %>% dplyr::summarise(x = median(pos))

  mycols <- rep(cols, length(chrs))[1:length(unique(chrs))]

  if(length(unique(chrs)) == 1){
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      xlab(paste('Chromosome', unique(chrs)))
  } else {
    plot <- ggplot(d, aes(x = pos, y = logp, color = CHR)) +
      scale_x_continuous(name = 'Chromosome', breaks = ticks$x, labels = ticks$CHR)
  }
  upper <- ceiling(max(d$logp))
  if(!is.null(ymax) && is.numeric(ymax)){
    upper <- ceiling(ymax)
  }
  plot <- plot + geom_point() + scale_color_manual(values = mycols) +
    scale_y_continuous(breaks = seq(1, upper, by = floor(upper/6))) +
    ylab(expression(-log[10](italic(p)))) +
    ggtitle(title) +
    theme_bw() +
    theme(axis.text.y = element_text(size = rel(2), color = 'black'),
          axis.text.x = element_text(size = rel(1.5), color = 'black'),
          legend.position = 'none', plot.title = element_text(size = rel(5), hjust = 0.5),
          panel.grid.minor = element_blank())

  if(annotate){
	 if(annotate.fld=='SNP'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- SNPlist[[1]]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- SNPlist[[2]]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size = 5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}else if(annotate.fld == 'Gene'){
		if(length(SNPlist)==1){
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=c(5,ymax))
		}else{
		d$labl1 <- ''
		d$labl1[match(SNPlist[[1]], d$SNP)] <- d$Gene[match(SNPlist[[1]], d$SNP)]
		d$labl2 <- ''
		d$labl2[match(SNPlist[[2]], d$SNP)] <- d$Gene[match(SNPlist[[2]], d$SNP)]
		plot <- plot +
		geom_text_repel(aes(label=d$labl1), size =5, color='black',ylim=lab2lim,min.segment.length=0.1)+
		geom_text_repel(aes(label=d$labl2), size = 5, color='orange',ylim=lab3lim,segment.alpha=0.4,min.segment.length=0.1)
		}
	}
  }

  if (!is.null(suggestiveline))
    plot <- plot + geom_hline(yintercept = suggestiveline, colour="brown", alpha = 0.33)
  if (!is.null(genomewideline))
    plot <- plot + geom_hline(yintercept = genomewideline, colour = "brown")
  if(!is.null(ymax) && is.numeric(ymax)){
    if(ymax <= max(d$logp) & ymax >= 0)
      plot <- plot + ylim(0, ymax)
  }
  plot
}


```


```{r}
SNPlist <- as.list(DMPs$Name)
```

```{r}
Soil5yr_manhattan <- DMPs %>% dplyr::select(chr, pos, P.Value, Name, GencodeBasicV12_NAME) 

# Renaming columns in Soil5yr_manhattan to match the function's expected names
names(Soil5yr_manhattan) <- c('CHR', 'BP', 'P', 'SNP', 'Gene')

# Remove "chr" prefix and convert to numeric
Soil5yr_manhattan$CHR <- as.numeric(sub("chr", "", Soil5yr_manhattan$CHR))
```

```{r}
# Running gg.manhattan with the modified Soil5yr_manhattan dataframe
gg.manhattan(Soil5yr_manhattan, chr.fld = 'CHR', pos.fld = 'BP', p.fld = 'P', snp.fld = 'SNP', gene.fld = 'Gene', annotate = FALSE, SNPlist = SNPlist)
```




QQ plot
```{r}
qq(DMPs$P.Value)
```



### DMR Analysis
```{r}
myAnnotation <- cpg.annotate(object = mVals, datatype = "array", what = "M", 
                             analysis.type = "differential", design = design, 
                             contrasts = FALSE, 
                             coef = "Soil5yrPreSamp_cont", arraytype = "EPIC")
str(myAnnotation)
```


```{r}
DMRs <- dmrcate(myAnnotation, lambda=1000, C=2)
results.ranges_Soil5yr_adj <- extractRanges(DMRs)
results.ranges_Soil5yr_adj
```


### Gene Ontology Adjusted Analysis
```{r}
enrichment_GO_Soil5yr_adj <- goregion(results.ranges_Soil5yr_adj[1:1], all.cpg = rownames(mVals), collection = "GO", array.type = "EPIC")
enrichment_GO_Soil5yr_adj <- enrichment_GO_Soil5yr_adj[order(enrichment_GO_Soil5yr_adj$P.DE),]
head(as.matrix(enrichment_GO_Soil5yr_adj), 10)
```

### Write DMR Results to .csv
```{r, eval=F}
results.ranges_Soil5yr_adj <- as.data.frame(results.ranges_Soil5yr_adj)

write.table(results.ranges_Soil5yr_adj, file="Simmons_ContinuousSoil5yrSamp_DMRs_CellDeconvAx_2024_07_18.csv", sep=",", row.names=FALSE)
```
